Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-07-05 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC_bak.tex
updates of files

\medskip
Description of physics and mechanisms of multiphase porous media flows were reported by \citet{wooding_1976} with focus on capillary pressure and flow regimes. A comprehensive review of force balances at the interface between immiscible fluids and resulting mechanisms for flow instabilities can be found in \citet{homsy_1987}. Flow instabilities can be classified as macroscopic and microscopic, the former is due to imposed boundary conditions, whereas the later is associated to local phenomena at the fluids interface \citep[\eg Kelvin-Helmholtz and Saffman-Taylor instabilities,][]{saffman_1959}. \blue{This work focuses on flow instabilities in two-phase systems due to viscous and stress forces (often referred as viscous instabilities or viscous fingering, Section \ref{section:ViscousInstabilities})}.

@@

\medskip 
Multi-fluid flow dynamics in porous media are described by continuity and momentum (extended Darcy's law) conservative equations for each fluid (or phases) with coupling mass terms (\ie density and saturation) appearing in both sets of equations. %Advanced numerical methods are in continuous development to accurately represent force balances in Darcean flows. 
Finite difference methods (FDM) have been extensively used in most industry-standard reservoir simulators~\citep{aziz_1986, chen_2005, chang_1990} with relative success. However, they are often limited to relatively simple geometries representing idealised geological formations through structured quadrilateral (2-D) and hexahedral (3-D) grid cells~\citep{mlacnik_2004, king_1999}. Additionally, FDM schemes often result in excessive numerical dispersion when strong heterogeneity (represented by permeability and porosity fields) is present~\citep{chavent_1986}.

\medskip
The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to be more efficient than FDM to solve fluid flow dynamics in complex geometries. %FEM discretisation is based upon piecewise representation of the solution space and the computational domain is divided into smaller subdomains (elements) where the solution in each element is constructed from basis functions. %Advantages of FEM over FVM are the better description of the geometry and complex shapes, better treatment of fluid flow and more accurate field-solution calculations using a smaller number of nodes, less memory space and shorter run times. %At this point let us remind that a comparison of computaional methods is difficult and depends always on the test-case.
%The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to be more efficient compared with FDM schemes when they are used to solve complex geometrical domains. FEM discretization is based upon a piecewise representation of the solution in terms of specified basis functions. The computational domain is divided into smaller domains (elements) and the solution in each element is constructed from the basis functions. Advantages of FEM over FVM are the better description of the geometry and complex shapes, better treatment of fluid flow and more accurate field-solution calculations using a smaller number of nodes, less memory space and shorter run times. 
Among FEM-based formulations for porous media, the control volume finite element methods~\citep[CVFEM,][]{fung_1992} has been widely used as it can guarantee local mass conservation and high-order numerical accuracy as well as being able to use tetrahedral geometry-conforming elements. In traditional CVFEM formulations, pressure and velocity are interpolated using piecewise linear FE basis functions, %$\underline{\bf v}$ and $\underline{\bf p}$)
while material properties and flow conditions (\eg phase saturation, density, temperature, species concentration etc) are represented with CV basis functions~\citep{voller_2009}.

%Saturation equations are solved explicitly after solving for the dual pressure-velocity at each non-linear iteration~\citep[a detailed description of the implicit pressure explicit saturation, IMPES, formulation can be found in][]{chen_2006,lux_2012}.

%%%%%%%%%%%%%%%
%\begin{comment}
%\begin{equation}
% p^{n} = \sum_{j} \hat{p_{j}} N_{\alpha} 
%\label{cvfe_pressure} 
%\end{equation}

%\begin{equation}
% S^{n} = \sum_{j} \hat{p_{j}} M_{b} 
%\label{cvfe_saturation}
%\end{equation} 
%\end{comment}
%%%%%%%%%%%%%%%

\medskip
Since geometries are captured by finite elements, constructed control volumes typically extend on each side of the interface which may have different properties. Therefore, some average values of the coupled velocity-pressure fields (defined in the FE space but projected onto the CV space) are applied across the CVs at these interfaces. These often lead to excessive numerical dispersion especially in highly heterogeneous media (represented by spatial-dependent permeability and porosity fields). In order to overcome such artificial numerical dispersion, a discontinuous hybrid finite element finite volume method (DFEFVM) formulation was introduced by \citet{nick_2011b, nick_2011a}. This novel discretisation scheme was designed to simulate flows through discrete fractured rocks in which CVs are divided along the interfaces of different materials. \citet{cumming_2011} demonstrated that CVFEM discretisation could also be used to solve Richards' equations (coupled mass conservation and Darcy equation) in heterogeneous media with relatively small computational overhead~\citep[compared with traditional coupled velocity-pressure based formulations, see also][]{cumming_phd2012}. %Fluxes over CVs were calculated based upon material properties, whereas the saturation field was volume-averaged at the interface of the materials, enforcing mass balance as described by \citet{kirkland_1992}.

%The conservation of mass along with Darcy's law are the governing equations while variables such as pressure, velocity,saturation and permeability define the behaviour of the flow. Finite difference methods have been applied to discretise the governing equations \citep{Luo2016,Moortgat2016,Hoteit2008} .
%\KCnote{in our case only - In conventional reservoirs the principal forces that affect fluid flow include, viscous forces, effects of gravity and permeability diffrences while in unconventional gas reservoirs flow simulation models take into account the phenomena of adsorption and diffusion \citep{abdus_2015}. Even though the mechanisms of the instabilities may vary, however it is still a function of relative mobility, gravity, viscous forces and permeability differences. All these parameters are represented under one term in the global mass balance equation.}

\medskip 
In this work, a novel CVFEM formulation, previously introduced by \citet{gomes_2017} (see also \citet{jackson_2015,salinas2015}), is used to numerically investigate formation and growth of viscous fingers in heterogeneous porous media. The continuity equation is embedded into the pressure equation to enforce mass conservation whilst ensuring that the force balance is preserved. A hybrid family of \PN[n]{m} triangular and tetrahedral FE pairs is used to discretise velocity and pressure. %, in which velocity is represented by $n$ th-order polynomials that are discontinuous across elements whereas pressure is represented by $m$ th-order polynomials that are continuous across the elements.
A sketch of the \PN[1]{2} FE-pair is shown in Fig.~\ref{fig:fem_cv}, in which velocity is represented by discontinuous and piecewise linear basis functions whereas pressure is interpolated through continuous and piecewise quadratic basis functions. Scalar fields are stored in CV space (Fig.~\ref{fig:fem_elem}). %The dual pressure and velocity fields are represented simultaneously (through non-linear projections) in FE and CV spaces.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Model Formulation}\label{equations_scheme}      
%Multiphase flow in porous media is described by mass conversation and Darcy equations for all the phase $\alpha$ ($\alpha$ = g, o, w, for gas, oil, and water phases). Different domains will produce different flow instabilities. In this paper, the test-cases are based on the discontinuous Galerkin (DG) method. DG-methods combine advantages of classical finite volume and finite element methods.

%\subsection{Numerical Formulation}
%%%%%%%%%%%%%%%
%\begin{comment}
%Flows in porous media are expressed by the Darcy's law and the saturation equation. At this stage the control volume finite element method is used (CVFEM). Darcy's law for immiscible multi-phase flow may be written in the form:
%\begin{equation}\label{e:darcy_eq}
%  \mathbf{q}_{\alpha} = -\frac{\mathcal{K}_{{r}_\alpha}\mathbf{K}}{\mu_{\alpha}}\left(\nabla p_{\alpha} - {\mathbf{s}_{u}}_{\alpha} \right),
%\end{equation}
%where $\mathbf{q}_{\alpha}$ is the $\alpha$-th phase Darcy velocity, $\mathbf{K}$ is the absolute permeability tensor of the porous medium, $\mathcal{K}_{{r}_\alpha}\left(S_{\alpha}\right)$ is the phase relative permeability, which is a function of the phase saturation $S_{\alpha}\left(\mathbf{r},t\right)$. $\mu_{\alpha}$, $p_{\alpha}$, $\rho_{\alpha}$, and $\mathbf{s}_{{u}_\alpha}$ are the phase dynamic viscosity, pressure, density and source term, which may include gravity and/or capillarity, respectively.

%Substituting the saturation-weighted Darcy velocity by $\mathbf{u}_\alpha= \mathbf{q}_\alpha/S_\alpha$, then Eqn.~\ref{e:darcy_eq} may be rewritten as:
%\begin{equation}
%  \mathbf{v}_\alpha={\underline {\underline \sigma}}_{\alpha} \mathbf{u}_{\alpha} = - \nabla p_{\alpha} + {\mathbf{s}_{u}}_{\alpha},
%  \label{e:force_bal}
%\end{equation}
%where ${\underline {\underline \sigma}}_{\alpha}=\mu_\alpha S_\alpha \left(\mathcal{K}_{{r}_\alpha}\mathbf{K}\right)^{-1}$ represents the implicit linearisation of the viscous frictional forces. The force per unit volume $\mathbf{v}_\alpha$, defined as ${\underline {\underline \sigma}}_{\alpha} \mathbf{u}_\alpha$, is used as a prognostic variable in this approach \citep{pavlidis_2014}.

%\noindent In matrix form, Eq.~\ref{e:force_bal} is: %~\ref{force-semi-disc} is: 
%\begin{equation}
%  {\mathbf M} \underline {\mathbf v} = -{\mathbf C} \underline{\mathbf p} + \underline {\mathbf s}_{u}, \label{force-balance-matrix-form}
%\end{equation}
%where $\underline{\bf v}$ and $\underline{\bf p}$ solution-vectors (also known as shape functions) are defined as:
%\begin{eqnarray}
%    &&\underline{{\bf v}} = \left( \left( {v_x},{v_y},{v_z} \right)_{1,1}, \left( {v_x},{v_y},{v_z} \right)  _{2,1},\ldots, \left( {v_x},{v_y},{v_z} \right)_{\mathcal{N}_\alpha,\mathcal{N}_u} \right)^{T} \;\text{and} \nonumber \\ 
%    && \underline{{\bf p}} = \left(p_1, p_2, p_3, \cdots, p_{\mathcal{N}_p}\right)^T. \nonumber% \nonumber
%\label{v_and_p}
%\end{eqnarray}

%The saturations are computed in CV space, whereas the permeability tensor $\left(\mathbf{K}\right)$ is assumed piece-wise constant in FE space. Thus the viscous-friction damping tensor $ {\underline {\underline \sigma}}_{\alpha}$ is piece-wise constant within the sub-control volumes defined through intersections between the elements and control volumes.
%Introducing a saturation-weighted Darcy velocity defined as $\mathbf{u}_\alpha= \mathbf{q}_\alpha/S_\alpha$, then Eqn.~\ref{e:darcy_eq} may be rewritten as:
%\begin{equation}
%  \mathbf{v}_\alpha={\underline {\underline \sigma}}_{\alpha} \mathbf{u}_{\alpha} = - \nabla p_{\alpha} + {\mathbf{s}_{u}}_{\alpha},
%  \label{e:force_balance}
%\end{equation}

%\noindent The ${\underline {\underline \sigma}}_{\alpha}=\mu_\alpha S_\alpha \left(\mathcal{K}_{{r}_\alpha}\mathbf{K}\right)^{-1}$ represents the implicit linearisation of the viscous frictional forces. Assuming a porous media domain $\Omega\subset\mathcal{R}^{n}$ ($n$ is the number of dimensions). 
%\end{comment}

%%%%%%%%%%%%%%%
%\begin{comment}
%The displacing and displaced fluids are considered to be incompressible and immiscible under isothermal and no source or sink conditions. In addition, we neglect the effects of capillarity and gravity in our simulations. Therefore the governing equations of conservation of mass and Darcy's law can be described below as,

%\begin{equation}
%  \mathbf{q}_{\alpha} = -\frac{\mathcal{K}_{{r}_\alpha}\mathbf{K}}{\mu_{\alpha}}\left(\nabla p_{\alpha} - {\mathbf{s}_{u}}_{\alpha} \right),
%\label{e:darcy_eq}
%\end{equation}
%where $\mathbf{q}_{\alpha}$ is the $\alpha$-th phase Darcy velocity, $\mathbf{K}$ is the absolute permeability tensor of the porous medium, $\mathcal{K}_{{r}_\alpha}\left(S_{\alpha}\right)$ is the phase relative permeability, which is a function of the phase saturation $S_{\alpha}\left(\mathbf{r},t\right)$. $\mu_{\alpha}$, $p_{\alpha}$, $\rho_{\alpha}$, and $\mathbf{s}_{{u}_\alpha}$ are the phase dynamic viscosity, pressure, density and source term, which may include gravity and/or capillarity, respectively.

%\begin{equation}
%  \mathbf{q}_{\alpha} = -\frac{\mathcal{K}_{{r}_\alpha}\left(S_{\alpha}\right)\mathbf{K}}{\mu_{\alpha}}\left( \nabla p_{\alpha} - {\mathbf{s}_{u}}_{\alpha} \right),
%S_{k}\left(\lambda_{k}\mathcal{K}\right)^{-1}\phi v_{\alpha} = \underline{\underline{\sigma}}_{\alpha} {\bf u}_{\alpha} = -\nabla p + \mathcal{S}_{u},
%\label{E:darcy_eqn}
%\end{equation}
%where $\mathbf{q}_{\alpha}$ is the $k\left(\in\left\{1,N_{p}\right\}\right)$-phase Darcy flow rate. $S$, $\mathbf{K}$, $\mathcal{K}_{r\alpha}\left(S_{\alpha}\right)$ and $\mu$ are saturation, absolute permeability tensor, phase relative permeability and viscosity, respectively. $s_{u}$ is a source term (e.g., gravity force) associated with the force balance and $p$ is the pressure.

% Defining the advective velocity averaged over the entire domain -- $\mathbf{u}_{\alpha}= \mathbf{q}_{\alpha}/S_{\alpha}$, then we may rewrite Eq.(\ref{E:darcy_ext}),
%\begin{equation}
%  {\underline {\underline \sigma}}_{\alpha} \mathbf{u}_{\alpha} = - \nabla p + {\mathbf{s}_{u}}_{\alpha},
%  \label{e:darcy_ext}
%\end{equation}
%$\underline{\underline{\sigma}}_{\alpha}=\mu_{\alpha}S_{\alpha}\left(\mathcal{K}_{r\alpha}\mathbf{K}\right)^{-1}$ represents the implicit linearisation of the viscous frictional forces, and is piecewise constant within each FE and is obtained via basis functions local to each CV within each element \citep{Jackson13,Radunz14}.
%\end{comment}
%%%%%%%%%%%%%%%

@@

\begin{eqnarray}
\left(\displaystyle\frac{\mu_{\alpha}S_{\alpha}}{{\mathbf K}\mathcal{K}_{r\alpha}}\right) {\mathbf u}_{\alpha} = \underline{\underline{\sigma}}_{\alpha} {\mathbf u}_{\alpha} = -\nabla p + \mathcal{S}_{u,\alpha}\hspace{1cm} \label{eqn:darcy_eqn} \\ %\text{ with } x_{i}\in\Omega,\; t>0, \label{eqn:darcy_eqn} \\
\phi\displaystyle\frac{\partial S_{\alpha} }{\partial t} +   \nabla \cdot \left( {\mathbf u}_{\alpha}  S_{\alpha}\right) =  \mathcal{S}_{cty,\alpha}\hspace{1cm} \text{ with } x_{i}\in\Omega,\; t>0,\label{eqn:saturation_eqn}
\end{eqnarray}

\noindent respectively, where $\mu$, ${\bf K}$, $p$ and $\phi$ are viscosity, absolute permeability, pressure and porosity, respectively. $\mathcal{S}_{u,\alpha}$ and $\mathcal{S}_{cty,\alpha}$ are source terms of the $\alpha$-phase and $\mathcal{K}_{r,\alpha}$ is the relative permeability. $\mathcal{S}_{u, \alpha}$ is the source term 
which may include gravity and/or capillarity and is related with momentum while the $\mathcal{S}_{cty,\alpha}$ is the source term related with mass (continuity equation). $S_{u, \alpha}$ is the source term for momentum while the saturation of the $\alpha$-phase with mass conservation constraints of $\sum\limits_{\alpha=1}^{\mathcal{N}_{p}} S_{\alpha} = 1$, where $\mathcal{N}_{p}$ denotes the number of phases. $\underline{\underline{\sigma}}_{\alpha}$ is an absorption-like term that represents the implicit linearisation of the viscous frictional forces.
%\begin{displaymath}
%\sum\limits_{\alpha=1}^{\mathcal{N}_{p}} S_{\alpha} = 1, 
%\end{displaymath}
%where $\mathcal{N}_{p}$ denotes the number of phases.

@@

\medskip
The formulation introduced here is applicable to $\mathcal{N}_{p}$ fluid phases and is based on two families of FE-pairs: \PNDG[n]{m} and \PN[n]{m}~\citep{cotter_2009a}, consistent with the dual pressure-velocity representation in CV space. In these families of FE-pairs, velocity is represented by $n^{\text{th}}$-order polynomials that are discontinuous across elements, whereas pressure is represented by $m^{\text{th}}$-order polynomials that may be either continuous or discontinuous (thus the notation \PN[n]{m} and \PNDG[n]{m}, respectively) across elements. Mass balance (continuity) equations are solved in CV space and a Petrov-Galerkin FEM is used to obtain high-order fluxes on CV boundaries, which are limited to yield bounded fields (\eg positive densities, saturations bounded between 0 and 1, etc). Simulations performed for this work were conducted using two types of elements: \PN[1]{1} and \PN[1]{2}.

@@

\medskip
Whilst saturation (and all saturation-dependent material properties such as relative permeability and capillary pressure) is calculated in CV space, absolute permeability is assumed piecewise constant in FE space. Saturation equations (Eqn.~\ref{eqn:saturation_eqn}) are discretised in space with CV basis function, $M_{i}$, and with the $\theta$-method in time~\citep{gomes_book_2012}. 
%Velocities across CV interfaces (within and between elements) are calculated through a directional-weighted flux-limited scheme based on upwind value of $\sigma$ at individual CV as described by \citet{jackson_2013}. 
Summing the discretised Eqn.~\ref{eqn:saturation_eqn} over all phases yields the global mass balance equation,
\begin{eqnarray}
 && \sum_{\alpha=1}^{N_{p}} \left\{\int\limits_{\Omega_{CVi}} \; M_{i} \; \frac{\phi\left(S_{\alpha i}^{n+1}-S_{\alpha i}^{n}\right)}{\Delta t} dV\right.  + \nonumber\\
 &&  \oint_{\Gamma_{CVi}} \left[\; \theta^{n+1/2}\; {\mathbf n}\cdot {\mathbf u}_{\alpha}^{n+1}S_{\alpha}^{n+1} \; + \; \left(1-\theta^{n+1/2}\right) \; {\mathbf u}_{\alpha}^{n}S_{\alpha}^{n}\right] d\Gamma \;- \nonumber\\
 &&  \left.\int_{\Omega_{CVi}} M_{i} \; \mathcal{S}_{cty,\alpha}^{n+\theta} \; dV\;\right\} =0,
\label{global_mass_balance}
\end{eqnarray}

@@

%\medskip
%The discretised global mass and force balance equations are solved using a multigrid-like approach as described by \citet{pavlidis2016} \citep[see also][]{salinas_2016}. The numerical formulation is fully described by \citet{gomes_2017} \citep[see also][]{salinas2015,adam_2016}.

@@

\medskip
In this work, all test-cases were performed in idealised geometries discretised with unstructured triangular and tetrahedral mesh using the \PN[1]{2} and \PN[1]{1} FE-pairs (Fig.~\ref{fig:fem_elem}). An implicit Crank-Nicolson time-stepping scheme was used with {\it a posteriori} adaptive time-step size targeting a maximum Courant-Friedrichs-Lewy condition \citep[CFL,][]{courant_1941} of 2. In most simulations, the domain was initially fully saturated with non-wetting fluid which was displaced by a pure (wetting) fluid at a prescribed initial velocity $\left(u^{\circ}\right)$. For simplicity, the porosity ($\phi$) of the domain was kept constant at 0.2 in all simulations, whereas the absolute permeability ($\mathbf{K}$) varied in space, \ie $\mathbf{K}=\mathbf{K}\left(x_{i}\right)$. Fluids are assumed incompressible and gravity was neglected. Initial set-up for the numerical simulations is summarised in Table \ref{table:setup}.

%%%%%%%%%%%% 
\subsection{Initial Model Validation}\label{section:results_initial_model_validation}
Numerical simulations \citep[based on lab experiments due to][]{evans_1994,dawe_2008} were conducted to demonstrate the model's capability to capture viscous crossflow during immiscible displacement in heterogeneous porous media. The 2D domain, shown in Fig.~\ref{fem_cv_represent_a}, is $4$ $\times$ $2$ unit-length and fully saturated with fluid 2 (VR=1). Fluid 1 is injected from the left-hand side of the domain with constant velocity of $u=1$. No normal flux boundary conditions are implied across the upper and lower borders. The domain consists of four regions in which each quarter is represented by a permeability value, Fig.~\ref{fem_cv_represent_a}(a). The spatial permeability distribution creates a rough pressure field, \ie pressure gradient is discontinuous across the interface between different permeability zones, therefore the fluid flows to the larger permeability region. %Figure \ref{fem_cv_represent_a}(b-c)describe the saturation field at different times.

>>>>

\medskip
Description of physics and mechanisms of multiphase porous media flows were reported by \citet{wooding_1976} with focus on capillary pressure and flow regimes. A comprehensive review of force balances at the interface between immiscible fluids and resulting mechanisms for flow instabilities can be found in \citet{homsy_1987}. Flow instabilities can be classified as macroscopic and microscopic, the former is due to imposed boundary conditions, whereas the later is associated to local phenomena at the fluids interface \citep[\eg Kelvin-Helmholtz and Saffman-Taylor instabilities,][]{saffman_1959}. This work focuses on flow instabilities in two-phase systems due to viscous and stress forces, often referred as viscous instabilities or viscous fingering (Section \ref{section:ViscousInstabilities}).

@@

\medskip 
Multi-fluid flow dynamics in porous media are described by continuity and momentum conservative equations for each fluid (or phases) with coupling mass terms (\ie density and saturation) appearing in both sets of equations. Finite difference methods (FDM) have been extensively used in most industry-standard reservoir simulators however, they are often limited to relatively simple geometries and often lead to excessive numerical dispersion when strong heterogeneity is present~\citep{chavent_1986}.
%representing idealised geological formations through structured quadrilateral (2-D) and hexahedral (3-D) grid cells~\citep{mlacnik_2004, king_1999}. Additionally, FDM schemes often result in excessive numerical dispersion when strong heterogeneity (represented by permeability and porosity fields) is present~\citep{chavent_1986}.

The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to be more efficient than FDM to solve fluid flow dynamics in complex geometries. %FEM discretisation is based upon piecewise representation of the solution space and the computational domain is divided into smaller subdomains (elements) where the solution in each element is constructed from basis functions. %Advantages of FEM over FVM are the better description of the geometry and complex shapes, better treatment of fluid flow and more accurate field-solution calculations using a smaller number of nodes, less memory space and shorter run times. %At this point let us remind that a comparison of computaional methods is difficult and depends always on the test-case.
%The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to be more efficient compared with FDM schemes when they are used to solve complex geometrical domains. FEM discretization is based upon a piecewise representation of the solution in terms of specified basis functions. The computational domain is divided into smaller domains (elements) and the solution in each element is constructed from the basis functions. Advantages of FEM over FVM are the better description of the geometry and complex shapes, better treatment of fluid flow and more accurate field-solution calculations using a smaller number of nodes, less memory space and shorter run times. 
%Among FEM-based formulations for porous media flows, the control volume finite element methods~\citep[CVFEM,][]{fung_1992} has been widely used as it can guarantee local mass conservation and high-order numerical accuracy as well as being able to use tetrahedral geometry-conforming elements. In traditional CVFEM formulations, pressure and velocity are interpolated using piecewise linear FE basis functions, %$\underline{\bf v}$ and $\underline{\bf p}$)
%while material properties and flow conditions (\eg phase saturation, density, temperature, species concentration etc) are represented with CV basis functions~\citep{voller_2009}. 
Among FEM-based formulations for porous media flows, the control volume finite element methods~\citep[CVFEM,][]{fung_1992} was designed to guarantee local mass conservation and high-order numerical accuracy as well as being able to use tetrahedral geometry-conforming elements. In traditional CVFEM formulations, pressure and velocity are interpolated using piecewise linear FE basis functions, while material properties and flow conditions (\eg phase saturation, density, temperature, species concentration etc) are represented with CV basis functions~\citep{voller_2009}.

Since geometries are captured by finite elements, constructed control volumes typically extend on each side of the interface which may have different properties. Therefore, average values of coupled velocity-pressure variables are applied across the CVs at the interfaces. This often leads to excessive numerical dispersion especially in highly heterogeneous media. In order to reduce artificial numerical dispersion associated to the upwind weighting of the advected quantity in regions of strong heterogeneity, a discontinuous hybrid finite element finite volume method (DFEFVM) formulation was introduced by \citet{nick_2011b, nick_2011a}. This novel discretisation scheme was designed to simulate flows through discrete fractured rocks in which CVs are divided along the interfaces of different materials. A control-volume distributed (CVD) scheme coupled with mixed FEM (MFEM) was introduced by \citet{edwards_2006}, where flow and rock variables are discretised in CV space whereas pressure and velocity are solved in FEM space. CVD flux-limiting scheme was coupled with high-order convection scheme to assess reduction of anisotropic numerical dispersion in flows through semi-impervious barrier (\ie a region of very low permeability).

% \citet{cumming_2011} demonstrated that CVFEM discretisation could also be used to solve Richards' equations (coupled mass conservation and Darcy equation) in heterogeneous media with relatively small computational overhead~\citep[compared with traditional coupled velocity-pressure based formulations, see also][]{cumming_phd2012}. %Fluxes over CVs were calculated based upon material properties, whereas the saturation field was volume-averaged at the interface of the materials, enforcing mass balance as described by \citet{kirkland_1992}.

\medskip 
In this work, a novel CVFEM formulation, previously introduced by \citet{gomes_2017} \citep[see also][]{jackson_2015,salinas2015}, is used to simulate viscous flow instabilities in porous media. The continuity equation is embedded into the pressure equation to enforce mass conservation whilst ensuring that the force balance is preserved. A hybrid family of \PN[n]{m} and \PNDG[n]{m} triangular and tetrahedral FE pairs is used to discretise velocity and pressure variables. Thus, the main aim of this work is to numerically investigate formation and growth of viscous fingers in heterogeneous porous media through a novel computational multi-fluid dynamics model (embedded in the next-generation flow simulator Fluidity/IC-FERST software). The two main aspects of this work are: (a) qualitative and quantitative validation of the model (Section~\ref{section:results_homo_hete}) for viscous flow instabilities and, (b) impact of mesh resolution on capturing the onset of viscous fingers formation.

%, in which velocity is represented by $n$ th-order polynomials that are discontinuous across elements whereas pressure is represented by $m$ th-order polynomials that are continuous across the elements.
%A sketch of the \PN[1]{2} FE-pair is shown in Fig.~\ref{fig:fem_cv}, in which velocity is represented by discontinuous and piecewise linear basis functions whereas pressure is interpolated through continuous and piecewise quadratic basis functions. Scalar fields are stored in CV space (Fig.~\ref{fig:fem_elem}). %The dual pressure and velocity fields are represented simultaneously (through non-linear projections) in FE and CV spaces.

\medskip
A brief description of the numerical formulation and viscous flow instabilities are introduced in Sections~\ref{equations_scheme} and~\ref{section:ViscousInstabilities}, respectively. Model set up and results including initial model-benchmark are presented in Section~\ref{section:results}. Impact of viscosity ratio on the fingers formation is also included in this section. Finally, concluding remarks are presented in Section~\ref{Section:Conclusion}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Model Formulation}\label{equations_scheme}

@@

\begin{eqnarray}
&&\left(\displaystyle\frac{\mu_{\alpha}S_{\alpha}}{{\mathbf K}\mathcal{K}_{r\alpha}}\right) {\mathbf u}_{\alpha} = \underline{\underline{\sigma}}_{\alpha} {\mathbf u}_{\alpha} = -\nabla p + \mathcal{S}_{u,\alpha},\hspace{1cm} \label{eqn:darcy_eqn} \\ %\text{ with } x_{i}\in\Omega,\; t>0, \label{eqn:darcy_eqn} \\
&&\phi\displaystyle\frac{\partial S_{\alpha} }{\partial t} +   \nabla \cdot \left( {\mathbf u}_{\alpha}  S_{\alpha}\right) =  \mathcal{S}_{cty,\alpha},\label{eqn:saturation_eqn}
\end{eqnarray} 
with $x_{i}\in\Omega,\; t>0$. $\mu$, ${\bf K}$, $p$ and $\phi$ are viscosity, absolute permeability, pressure and porosity, respectively. $\mathbf{u}_{\alpha}$ is the saturation-weighted Darcy velocity of the $\alpha$-phase. $\mathcal{S}_{u,\alpha}$ and $\mathcal{S}_{cty,\alpha}$ are source terms of the $\alpha$-phase related to the momentum (\eg gravity, capillarity etc) and continuity (\eg geochemical reactions etc) equations. Finally, $\mathcal{K}_{r,\alpha}$ is the relative permeability, and $S_{\alpha}$ is the  $\alpha$-phase saturation with mass conservation constraints of $\sum\limits_{\alpha=1}^{\mathcal{N}_{p}} S_{\alpha} = 1$, where $\mathcal{N}_{p}$ denotes the number of phases. $\underline{\underline{\sigma}}_{\alpha}$ is an absorption-like term that represents the implicit linearisation of the viscous frictional forces.
%\begin{displaymath}
%\sum\limits_{\alpha=1}^{\mathcal{N}_{p}} S_{\alpha} = 1, 
%\end{displaymath}
%where $\mathcal{N}_{p}$ denotes the number of phases.

@@

\medskip
The formulation introduced here is applicable to $\mathcal{N}_{p}$ fluid phases and is based on two families of FE-pairs: \PNDG[n]{m} and \PN[n]{m}~\citep{cotter_2009a}, consistent with the dual pressure-velocity representation in CV space. In these families of FE-pairs, velocity is represented by $n^{\text{th}}$-order polynomials that are discontinuous across elements, whereas pressure is represented by $m^{\text{th}}$-order polynomials that may be either continuous or discontinuous (thus the notation \PN[n]{m} and \PNDG[n]{m}, respectively) across elements.

Mass balance (continuity) equations are solved in CV space and a Petrov-Galerkin FEM is used to obtain high-order fluxes on CV boundaries. Smooth transition between first- and high-order fluxes are enforced through an extrema detection scheme based on normalised variable diagram \citep[NVD,][]{jasak_1999,darwish_2003} applied across faces of neighbouring CVs. Resulting normalised upwind face values are used to calculate flux-limited solutions \citep[based on total variation diminishing criteria, TVD,][]{piperno_1998} to yield bounded fields \citep[\eg positive densities, saturations bounded between 0 and 1 etc. See][for more details on the flux-limiting scheme within the model]{gomes_book_2012}. Simulations performed for this work were conducted using three types of elements: \PN[1]{1}, \PN[1]{2} and \PNDG[1]{1} (Section~\ref{section:results_3D}).

@@

\medskip
Whilst saturation (and all saturation-dependent material properties such as relative permeability and capillary pressure) is calculated in CV space, absolute permeability is assumed piecewise constant in FE space. Saturation equations (Eqn.~\ref{eqn:saturation_eqn}) are discretised in space with CV basis function, $M_{i}$, and with the $\theta$-method in time~\citep{gomes_book_2012}. Velocities across CV interfaces (within and between elements) are calculated through a directional-weighted flux-limited scheme based on upwind value of $\underline{\underline{\sigma}}$ at individual CV as described by \citet{gomes_2017}. Summing the discretised Eqn.~\ref{eqn:saturation_eqn} over all phases yields the global mass balance equation,
\begin{eqnarray}
 && \sum_{\alpha=1}^{N_{p}} \left\{\int\limits_{\Omega_{CVi}} \; M_{i} \; \frac{\phi\left(S_{\alpha i}^{n+1}-S_{\alpha i}^{n}\right)}{\Delta t} dV\right.  + \nonumber\\
 &&  \oint_{\Gamma_{CVi}} \left[\; \theta^{n+1/2}\; {\mathbf n}\cdot {\mathbf u}_{\alpha}^{n+1}S_{\alpha}^{n+1} \; + \; \left(1-\theta^{n+1/2}\right) \; {\mathbf u}_{\alpha}^{n}S_{\alpha}^{n}\right] d\Gamma \;- \nonumber\\
 &&  \left.\int_{\Omega_{CVi}} M_{i} \; \mathcal{S}_{cty,\alpha}^{n+\theta} \; dV\;\right\} =0,
\label{global_mass_balance}
\end{eqnarray}

@@

%\medskip
The discretised global mass and force balance equations are solved using a multigrid-like approach. The numerical formulation is fully described by \citet{gomes_2017} \citep[see also][]{salinas2015,adam_2016}.

@@

\medskip
In this work, all test-cases were performed in idealised geometries discretised with unstructured triangular and tetrahedral mesh using the \PN[1]{2}, \PN[1]{1} (Fig.~\ref{fig:fem_elem}) and \PNDG[1]{1} (Section~\ref{section:results_3D})  FE-pairs. An implicit Crank-Nicolson time-stepping scheme was used with {\it a posteriori} adaptive time-step size targeting a maximum Courant-Friedrichs-Lewy condition \citep[CFL,][]{courant_1941} of 2. In most simulations, the domain was initially fully saturated with non-wetting fluid which was displaced by a pure (wetting) fluid at a prescribed initial velocity $\left(u^{\circ}\right)$. For simplicity, the porosity ($\phi$) of the domain was kept constant at 0.2 in all simulations, whereas the absolute permeability ($\mathbf{K}$) varied in space, \ie $\mathbf{K}=\mathbf{K}\left(x_{i}\right)$. Fluids are assumed incompressible and gravity was neglected. Initial set-up for the numerical simulations is summarised in Table \ref{table:setup}.

%%%%%%%%%%%% 
\subsection{Initial Model Validation}\label{section:results_initial_model_validation}
Numerical simulations \citep[based on lab experiments due to][]{evans_1994,dawe_2008} were conducted to demonstrate the model's capability to capture viscous crossflow during immiscible displacement in heterogeneous porous media. The 2D domain, shown in Fig.~\ref{fem_cv_represent_a}, is $4$ $\times$ $2$ unit-length and fully saturated with fluid 2 (VR=1). Fluid 1 is injected from the left-hand side of the domain with constant velocity of $u=1$. Boundary conditions also include no-flux across upper and lower borders. The domain consists of four regions in which each quarter is represented by a permeability value, Fig.~\ref{fem_cv_represent_a}(a). The spatial permeability distribution creates a rough pressure field, \ie pressure gradient is discontinuous across the interface between different permeability zones, therefore the fluid flows to the larger permeability region. %Figure \ref{fem_cv_represent_a}(b-c)describe the saturation field at different times.
